---
title: "Empirical Case Studies in Selection Analysis"
author: "Stroud Lab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical Case Studies in Selection Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Load required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
library(fields)
library(here)

# Load framework functions (Assuming execution from the project root)
fn_files <- list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE)
for (f in fn_files) source(f)

plot_files <- list.files(here("R", "plotting"), pattern = "\\.R$", full.names = TRUE)
for (f in plot_files) source(f)

script_files <- list.files(here("R", "scripts"), pattern = "\\.R$", full.names = TRUE)
script_files <- script_files[!grepl("0.0_initialize\\.R$", script_files)]
for (f in script_files) source(f)

# Helper function to find data files dynamically
find_data_file <- function(pattern) {
  files <- list.files(here(), pattern = pattern, recursive = TRUE, full.names = TRUE)
  if (length(files) > 0) return(files)
  return(NA)
}
```

## Introduction

This vignette demonstrates the complete quantitative genetic framework for analyzing natural selection on phenotypic traits, following the Lande & Arnold (1983) approach. We walk through three famous case studies using datasets representing fish, birds, and plants. 

These examples highlight how to handle both continuous and binary fitness metrics, single-year versus multi-year temporal analyses, and correlational fitness landscapes. By walking through these steps dynamically, you can explore the results directly in this document without having to run the individual test scripts manually.

---

## 1. Puppyfish Case Study: Morphology and Survival

In this first example, we examine survival (binary) and growth (continuous) in Crescent Pond puppyfish based on morphological traits such as jaw, eye, and body size.

### Data Preparation

First, we load and prepare the selection data. The `prepare_selection_data()` function automatically standardizes the traits (mean = 0, variance = 1) and handles missing data.

```{r fish-data}
crescent_path <- find_data_file("Crescent.*Pond.*\\.csv$")
crescent_data <- read.csv(crescent_path)

fitness_binary <- "survival"
fitness_continuous <- "ln.growth"
morphological_traits <- c("jaw", "eye", "body", "nasal", "mouth", "SL")

test_data <- crescent_data %>%
  select(all_of(c(fitness_binary, fitness_continuous, morphological_traits))) %>%
  filter(complete.cases(.))

prepared_binary <- prepare_selection_data(
  data = test_data,
  fitness_col = fitness_binary,
  trait_cols = morphological_traits,
  standardize = TRUE
)
```

### Estimating Selection Coefficients

We can estimate linear directional selection ($\beta$) and nonlinear stabilizing/disruptive selection ($\gamma$) in a single step using `selection_coefficients()`.

```{r fish-selection}
selection_binary <- selection_coefficients(
  data = prepared_binary,
  fitness_col = fitness_binary,
  trait_cols = morphological_traits,
  fitness_type = "binary",
  standardize = FALSE
)

knitr::kable(head(selection_binary), caption = "Top Selection Gradients for Puppyfish Survival")
```

### Visualizing Fitness Surfaces

We can visualize univariate and bivariate adaptive landscapes using penalized regression splines.

```{r fish-surface}
# Univariate fitness spline
jaw_spline <- univariate_spline(
  data = prepared_binary,
  fitness_col = fitness_binary,
  trait_col = "jaw",
  fitness_type = "binary",
  k = 6
)

plot_univariate_fitness(jaw_spline, trait_col = "jaw") +
  labs(title = "Fitness Surface: Jaw Size")

# Bivariate (correlational) fitness surface
tps_result <- correlated_fitness_surface(
  data = prepared_binary,
  fitness_col = fitness_binary,
  trait_cols = c("jaw", "eye"),
  method = "auto",
  grid_n = 20
)

plot_correlated_fitness(tps_result, trait_cols = c("jaw", "eye")) +
  labs(title = "Bivariate Fitness Surface: Jaw vs Eye")
```

---

## 2. Darwin's Finches Case Study: Temporal Analysis

Selection pressures often fluctuate across generations. Here, we analyze bird survival across multiple years, focusing on a principal component axis representing beak size (`Beak_PC1`).

### Loading and Processing Bird Data

We'll build a synthetic `Beak_PC1` from raw beak dimensions.

```{r bird-data}
bird_path <- find_data_file("bird_data\\.csv$")
bird_study <- read.csv(bird_path, stringsAsFactors = FALSE)
bird_study <- bird_study[, !grepl("^X\\.y\\.", names(bird_study))]

beak_cols <- c("MedianBeakLength", "MedianBeakWidth", "MedianBeakDepth")
if (all(beak_cols %in% names(bird_study))) {
  beak_complete <- bird_study[, beak_cols]
  ok <- complete.cases(beak_complete)
  
  if (sum(ok) > 10) {
    pca_fit <- prcomp(beak_complete[ok, ], scale. = TRUE)
    pc1_scores <- rep(NA_real_, nrow(bird_study))
    pc1_scores[ok] <- pca_fit$x[, 1]
    
    # Ensure Beak_PC1 positively correlates with depth
    cor_depth <- suppressWarnings(cor(pc1_scores[ok], beak_complete[ok, "MedianBeakDepth"]))
    if (!is.na(cor_depth) && cor_depth < 0) pc1_scores <- -pc1_scores
    
    bird_study$Beak_PC1 <- pc1_scores
  }
}
```

### Temporal Selection Analysis

We loop over multiple years of survival data to observe how the quadratic (disruptive/stabilizing) selection coefficient changes.

```{r bird-temporal}
# Helper to structure data correctly given paper definitions
prepare_basic_survival_data <- function(year, last_year = 2011) {
  current_col <- paste0("y.", year)
  later_cols  <- paste0("y.", (year + 1):last_year)
  
  if (!all(c(current_col, later_cols) %in% names(bird_study))) return(NULL)
  
  bird_study %>%
    filter(.data[[current_col]] == 1) %>%
    mutate(
      fitness = as.numeric(rowSums(across(all_of(later_cols)), na.rm = TRUE) > 0),
      year = year
    )
}

available_years <- 2003:2010
yearly_results <- list()

for (year in available_years) {
  year_data <- prepare_basic_survival_data(year)
  
  if (!is.null(year_data) && nrow(year_data) >= 20 && var(year_data$fitness, na.rm = TRUE) > 0) {
    # Single-trait disruptive selection test
    df <- year_data %>% filter(!is.na(Beak_PC1), !is.na(fitness))
    df$Beak_PC1 <- as.numeric(scale(df$Beak_PC1))
    
    m <- glm(fitness ~ Beak_PC1 + I(Beak_PC1^2), data = df, family = binomial)
    coefs <- summary(m)$coefficients
    
    if (nrow(coefs) >= 3) {
      yearly_results[[as.character(year)]] <- data.frame(
        Year = year,
        Sample_Size = nrow(df),
        Quadratic_Coefficient = coefs[3, 1],
        P_Value = coefs[3, 4],
        Significant = coefs[3, 4] < 0.05
      )
    }
  }
}

if (length(yearly_results) > 0) {
  yearly_summary <- bind_rows(yearly_results)
  knitr::kable(yearly_summary, caption = "Multi-Year Quadratic Selection Results")
  
  ggplot(yearly_summary, aes(x = Year, y = Quadratic_Coefficient, color = Significant)) +
    geom_point(aes(size = Sample_Size)) +
    geom_line(alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
    labs(
      title = "Temporal Variation in Disruptive Selection",
      subtitle = "Quadratic Selection on Beak Axis (Beak_PC1)",
      y = "Quadratic Selection Coefficient",
      size = "Sample Size"
    ) +
    scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red")) +
    theme_minimal()
}
```

---

## 3. Aster Plant Case Study: Correlational Selection

In our final example, we look at plant fecundity (a continuous fitness metric). We analyze multivariate selection gradients and temporal shifts for traits like Specific Leaf Area (`SLA`), carbon isotopes (`deltaC13`), snow-free days (`FDsnow`), and height.

### Structuring Plant Data

We assemble the trait and fitness matrices to capture continuous reproduction variables.

```{r plant-data}
# Find datasets
data1_path <- find_data_file("Aster_analyses_2011_Cohort\\.txt$")
data3_path <- find_data_file("Aster_analyses_2011_Cohort_full\\.txt$")

if (!is.na(data1_path) && !is.na(data3_path)) {
  data1 <- read.delim(data1_path, sep = "\t")
  data3 <- read.delim(data3_path, sep = "\t")
  
  c13_col <- grep("c13|deltaC13", names(data3), ignore.case = TRUE, value = TRUE)
  
  traits_2011 <- data3 %>%
    select(geno, elevation, season, SLA, !!c13_col, FDsnow, height) %>%
    rename(deltaC13 = !!c13_col) %>%
    distinct(geno, season, .keep_all = TRUE)
  
  fitness_wide <- data1 %>%
    select(geno, season, varb, resp) %>%
    filter(!is.na(resp)) %>%
    pivot_wider(
      names_from = c(season, varb),
      names_sep = "_",
      names_prefix = "Year",
      values_from = resp
    )
  
  analysis_data <- traits_2011 %>% left_join(fitness_wide, by = "geno")
  
  analysis_data_clean <- analysis_data %>%
    filter(!is.na(SLA), !is.na(deltaC13), !is.na(FDsnow), !is.na(height))
  
  names(analysis_data_clean) <- make.names(names(analysis_data_clean))
  trait_cols <- c("SLA", "deltaC13", "FDsnow", "height")
  
  main_fecund_col <- grep("^Year2013_.*fecund", names(analysis_data_clean), value = TRUE)
}
```

### Multivariate Selection Gradients

```{r plant-multivariate}
if (exists("analysis_data_clean")) {
  multi_trait_result <- selection_coefficients(
    data = analysis_data_clean,
    fitness_col = main_fecund_col,
    trait_cols = trait_cols,
    fitness_type = "continuous",
    standardize = TRUE
  )
  
  knitr::kable(multi_trait_result, caption = "Multivariate Selection Gradients (2013 Fecundity)")
}
```

## Summary

This framework provides a unified pipeline to:
1. Clean, standardize, and format survival and fecundity metrics.
2. Estimate linear gradients ($\boldsymbol{\beta}$) and nonlinear correlational/disruptive gradients ($\boldsymbol{\gamma}$).
3. Visualize adaptive landscapes effectively across taxonomic groups.