# Evolutionary Selection Analysis Framework

## 1. Overview

This project implements a comprehensive quantitative genetic framework for analyzing natural selection on phenotypic traits. The framework follows the Lande & Arnold (1983) approach while extending it with modern nonparametric methods for visualizing fitness surfaces and adaptive landscapes.

**Key capabilities:**

| Analysis Type | Description |
|---------------|-------------|
| Selection differentials ($\mathbf{S}$) | Total selection acting on traits, including indirect effects through correlated traits |
| Linear selection gradients ($\boldsymbol{\beta}$) | Direct directional selection on each trait, controlling for correlations with other traits |
| Nonlinear selection gradients ($\boldsymbol{\gamma}$) | Stabilizing/disruptive ($\gamma_{ii}$) and correlational ($\gamma_{ij}$) selection |
| Univariate correlated fitness functions | GAM spline-based fitness functions relating individual fitness to a single trait |
| Bivariate correlated fitness surfaces | GAM or thin-plate spline (TPS) surfaces relating individual fitness to two traits simultaneously |
| Adaptive landscapes | Population mean fitness as a function of population mean phenotype, estimated via simulation |
| Binary fitness handling | OLS for gradient estimation; logistic regression (GLM) for valid p-values |
| Group-level analyses | Stratified standardization and selection estimation within groups (e.g., year, site, population) |

---

## 2. Mathematical Framework

### 2.1 Relative Fitness

Individual fitness is converted to relative fitness to enable comparisons 
across populations and generations:

$$w_i = \frac{W_i}{\bar{W}}$$

where:
- $W_i$ = absolute fitness of individual $i$ (e.g., offspring count, survival status)
- $\bar{W} = \frac{1}{n}\sum_{i=1}^n W_i$ = mean population fitness
- $w_i$ = relative fitness of individual $i$

**Why relative fitness?** Selection operates on relative, not absolute, 
differences in fitness. This standardization ensures that selection gradients 
are comparable across studies.

**Group-level note:** When a grouping variable is specified (e.g., year, site), 
relative fitness is calculated separately within each group, so that individuals 
are compared only to others in the same context.

**File:** `1_prepare_selection_data.R`

### 2.2 Trait Standardization

Traits are standardized to mean 0 and variance 1 before analysis:

$$z_i = \frac{x_i - \bar{x}}{s_x}$$

where:
- $x_i$ = original trait value
- $\bar{x}$ = sample mean
- $s_x$ = sample standard deviation
- $z_i$ = standardized trait value

**Purpose:** Standardization puts all traits on the same scale, allowing direct comparison of selection gradients. A $\beta = 0.5$ means that a one-standard-deviation increase in the trait increases relative fitness by 0.5 units.
**File:** `1_prepare_selection_data.R`

### 2.3 Selection Differential

The selection differential ($S$) measures the total selection acting on a trait, including both direct and indirect effects through correlated traits:

$$S = \text{Cov}(z, w) = \mathbb{E}[(z - \bar{z})(w - \bar{w})]$$

When traits are standardized ($\bar{z} = 0$) and fitness is relativized ($\bar{w} = 1$), this simplifies to $$S = \frac{1}{n}\sum_{i=1}^n z_i w_i$$, which is the form used in this implementation.

For multiple traits, the vector of selection differentials is:
$$\mathbf{S} = \text{Cov}(\mathbf{z}, w) = \begin{bmatrix} \text{Cov}(z_1, w) \\ \text{Cov}(z_2, w) \\ \vdots \\ \text{Cov}(z_m, w) \end{bmatrix}$$

**Interpretation:** $S$ represents the shift in trait mean after one generation of selection. A positive $S$ indicates the trait mean increases after selection.
**File:** `selection_differential.R`

### 2.4 Linear Selection Gradients

Linear selection gradients ($\boldsymbol{\beta}$) measure the direct directional selection on each trait, controlling for correlations with other traits:

$$w = \alpha + \boldsymbol{\beta}^T \mathbf{z} + \varepsilon$$

In matrix form:

$$\boldsymbol{\beta} = \mathbf{P}^{-1} \mathbf{S}$$

where $\mathbf{P}$ is the phenotypic variance-covariance matrix.

For a model with two traits:

$$w = \alpha + \beta_1 z_1 + \beta_2 z_2 + \varepsilon$$

**Interpretation:** $\beta_i$ is the partial regression coefficient—the change in relative fitness for a one-standard-deviation increase in trait $i$, holding all other traits constant.
**File:** `analyze_linear_selection.R`

### 2.5 Nonlinear Selection Gradients

Quadratic selection gradients ($\boldsymbol{\gamma}$) measure nonlinear selection, including stabilizing/disruptive selection (diagonal elements) and correlational selection (off-diagonal elements):

$$w = \alpha + \boldsymbol{\beta}^T \mathbf{z} + \frac{1}{2} \mathbf{z}^T \boldsymbol{\gamma} \mathbf{z} + \varepsilon$$

For two traits, this expands to:

$$w = \alpha + \beta_1 z_1 + \beta_2 z_2 + \frac{1}{2}\gamma_{11} z_1^2 + \frac{1}{2}\gamma_{22} z_2^2 + \gamma_{12} z_1 z_2 + \varepsilon$$

**Important:** The $\frac{1}{2}$ factor in the quadratic terms follows Lande & Arnold's convention, making $\gamma_{ii}$ the second derivative of the fitness surface. In practice, most implementations (including this one) report $2 \times$ the quadratic coefficient from regression, which equals $\gamma_{ii}$.

**Interpretation:**

| $\gamma$ value | Type | Interpretation |
|----------------|------|----------------|
| $\gamma_{ii} < 0$ | Stabilizing selection | Intermediate trait values have highest fitness |
| $\gamma_{ii} > 0$ | Disruptive selection | Extreme trait values have highest fitness |
| $\gamma_{ij} > 0$ | Positive correlational | Selection favors positive correlation between traits |
| $\gamma_{ij} < 0$ | Negative correlational | Selection favors negative correlation between traits |
**File:** `analyze_nonlinear_selection.R`, `analyze_disruptive_selection.R`

### 2.6 Nonparametric Fitness Functions and Surfaces

Traditional quadratic models impose a fixed shape on the fitness function. Nonparametric methods allow the data to reveal unexpected patterns without assuming a particular functional form.

**Univariate fitness function (individual level):**

$$w = \alpha + s(z) + \varepsilon$$

where $s(z)$ is a smooth function estimated using penalized regression splines (GAM with REML smoothing parameter selection).

**Bivariate correlated fitness surface (individual level):**

$$w = \alpha + f(z_1, z_2) + \varepsilon$$

where $f(z_1, z_2)$ is a two-dimensional smooth surface estimated via:
- **GAM** (tensor product smooth): used for binary fitness
- **Thin-plate spline (TPS)**: used for continuous fitness

**GAM formulation:**
$$w \sim \text{family}(\eta), \quad \eta = \alpha + s(z_1, z_2)$$

Note: These surfaces describe **individual-level** fitness as a function of individual phenotype, they are correlated fitness surfaces, not adaptive landscapes. See Section 2.7 for the population-level adaptive landscape.

**File:** `correlated_fitness_surface.R`

### 2.7 Adaptive Landscapes (Population Level)

Following Sewall Wright's original concept, the adaptive landscape describes mean fitness as a function of population mean phenotype:

$$
\bar{W} = g(\bar{z}_1, \bar{z}_2)
$$

where:

- $\bar{W}$ = mean fitness of the population  
- $\bar{z}_1, \bar{z}_2$ = population mean phenotypes  

Unlike correlated fitness surfaces (which use individual data), adaptive landscapes require simulation because they represent a theoretical construct—the expected mean fitness for a population with a given mean phenotype.

**Calculation procedure:**

For each candidate population mean $\bar{\mathbf{z}}$:

**1. Simulate individuals around the mean**

$$
\mathbf{z}_{\text{ind}} \sim \mathcal{N}(\bar{\mathbf{z}}, \mathbf{P})
$$

where $\mathbf{P}$ is the within-population phenotypic variance–covariance matrix.

**2. Predict individual fitness**

$$
w_{\text{ind}} = \hat{f}(\mathbf{z}_{\text{ind}})
$$

**3. Compute mean fitness**

$$
\bar{W}(\bar{\mathbf{z}}) =
\frac{1}{N_{\text{sim}}}
\sum_{j=1}^{N_{\text{sim}}} w_{\text{ind}, j}
$$

The resulting surface:

$$
\bar{W} = g(\bar{z}_1, \bar{z}_2)
$$

is the **adaptive landscape**.

**Mathematical distinction**

| Surface | Level | Formula | Interpretation |
|--------|------|--------|----------------|
| Correlated Fitness | Individual | $w = f(\mathbf{z})$ | Current selection pattern |
| Adaptive Landscape | Population | $\bar{W} = g(\bar{\mathbf{z}})$ | Evolutionary potential |

**File:** `adaptive_landscape.R`

### 2.8 Model Selection and Comparison

The framework provides tools to compare correlated fitness surfaces with adaptive landscapes:

$$
\text{Comparison} = \text{Correlated Fitness}(\mathbf{z}) \; \text{vs} \; \text{Adaptive Landscape}(\bar{\mathbf{z}})
$$

This comparison highlights the difference between:

- **Empirical selection** (individual-level fitness)  
- **Theoretical evolutionary potential** (population-level mean fitness)  

**File:** `compare_fitness_surfaces.R`

---

## 3. Framework Structure

### 3.1 Core Scripts and Their Roles

| Script | Purpose | Key Functions | Mathematical Component |
|--------|---------|--------------|------------------------|
| `1_prepare_selection_data.R` | Data preprocessing | Standardization, relative fitness, NA handling, group support | $z_i = \frac{x_i - \bar{x}}{s_x}$, $w_i = \frac{W_i}{\bar{W}}$ |
| `detect_family.R` | Fitness type detection | Binary, count, proportion, or continuous | $W_i \in \{0,1\}$ vs $W_i \in \mathbb{Z}^+$ vs $W_i \in \mathbb{R}^+$ |
| `selection_differential.R` | Selection differential | $S = \text{Cov}(z, w)$, group-level support | $S = \frac{1}{n}\sum_i z_i w_i$ (when $\bar{z}=0$) |
| `selection_coefficients.R` | Main wrapper | Integrates linear, nonlinear, and extraction steps | $\boldsymbol{\beta}$, $\boldsymbol{\gamma}$, $\gamma_{ij}$ |
| `2_linear_selection_analysis.R` | Linear gradients | OLS for $\beta$; GLM for p-values (binary) | $w = \alpha + \boldsymbol{\beta}^T \mathbf{z} + \varepsilon$ |
| `3_nonlinear_selection_analysis.R` | Quadratic and correlational gradients | OLS for $\gamma$; GLM for p-values (binary); VIF check | $w = \alpha + \boldsymbol{\beta}^T\mathbf{z} + \frac{1}{2}\mathbf{z}^T\boldsymbol{\gamma}\mathbf{z} + \varepsilon$ |
| `4_disruptive_selection_analysis.R` | Single-trait nonlinear selection | Stabilizing/disruptive test; group support | $\gamma_{ii} = 2\hat{\beta}_{ii}$, $\text{SE}(\gamma_{ii}) = 2\,\text{SE}(\hat{\beta}_{ii})$ |
| `extract_results.R` | Coefficient extraction | Parses OLS and GLM outputs; handles binary/continuous | $\beta$, $\gamma_{ii}$, $\gamma_{ij}$ |
| `univariate_spline.R` | 1D correlated fitness function | GAM with REML; group support | $w = \alpha + s(z) + \varepsilon$ |
| `correlated_fitness_surface.R` | 2D correlated fitness surface | GAM (binary) or TPS (continuous); group support | $w = \alpha + f(z_1, z_2) + \varepsilon$ |
| `5_adaptive_landscape_analysis.R` | Adaptive landscape | Simulation via `mvrnorm`; positive-definite correction | $\bar{W}(\bar{\mathbf{z}}) = \frac{1}{N}\sum_j \hat{w}_j$, $\mathbf{z}_j \sim \mathcal{N}(\bar{\mathbf{z}}, \mathbf{P})$ |
| `compare_fitness_surfaces.R` | Surface comparison | Optimum distance; surface correlation via `akima` | $d = \|\mathbf{z}^* - \bar{\mathbf{z}}^*\|$, $r = \text{Cor}(\hat{f}, \bar{W})$ |
| `plot_correlated_fitness.R` | 2D fitness visualisation | Contour plots; enhanced version with original data overlay | $w = f(z_1, z_2)$ |
| `plot_univariate_fitness.R` | 1D fitness visualisation | Line plot with 95% CI bands | $\hat{w} \pm 1.96 \times \text{SE}$ |
| `plot_adaptive_landscape.R` | Adaptive landscape visualisation | 2D contour and 3D perspective view | $\bar{W}(\bar{\mathbf{z}})$ |
| `plot_fitness_surfaces_comparison.R` | Comparison visualisation | Side-by-side and overlay plots | Individual vs population optima |

---

## 4. Installation and Requirements

### 4.1 Required Packages, functions and scripts

### Required Packages

```{r, message=FALSE, warning=FALSE}
required_packages <- c(
  "mgcv", # GAM models for spline-based fitness functions
  "fields", # Thin-plate splines (Tps) for continuous fitness surfaces
  "MASS", # Multivariate normal simulation (mvrnorm) for adaptive landscape
  "Matrix", # Positive-definite matrix correction (nearPD)
  "car", # Type III ANOVA and variance inflation factors (VIF)
  "akima", # Surface interpolation for fitness surface comparison
  "ggplot2", # Plotting
  "dplyr", # Data manipulation and group-level operations
  "ggrepel", # Non-overlapping text labels in plots
  "viridis", # Perceptually uniform color scales
  "patchwork" # Combining multiple plots
)

missing <- required_packages[!required_packages %in% installed.packages()[, "Package"]]

if (length(missing) > 0) {
  install.packages(missing)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

### Required functions and scripts 
```{r, message=FALSE, warning=FALSE}
function_files <- c(
  "compare_fitness_surfaces.R",
  "correlated_fitness_surface.R",
  "detect_family.R",
  "extract_results.R",
  "extract_results.R",
  "selection_differential.R",
  "selection_coefficients.R",
  "univariate_spline.R"
)

scripts_files <- c(
  "1_prepare_selection_data.R",
  "2_linear_selection_analysis.R",
  "3_nonlinear_selection_analysis.R",
  "4_disruptive_selection_analysis.R",
  "5_adaptive_landscape_analysis.R"
)

plotting_files <- c(
  "plot_adaptive_landscape.R",
  "plot_correlated_fitness.R",
  "plot_univariate_fitness.R",
  "plot_fitness_surfaces_comparison.R"
)

all_files <- list(
  functions = list(path = "R/functions", files = function_files),
  scripts   = list(path = "R/scripts", files = scripts_files),
  plotting  = list(path = "R/plotting", files = plotting_files)
)

for (group in all_files) {
  for (f in group$files) {
    file_path <- file.path(group$path, f)

    if (file.exists(file_path)) {
      source(file_path)
      cat("Loaded:", file_path, "\n")
    } else {
      cat("Missing:", file_path, "\n")
    }
  }
}
```

---

## 5. Detailed Function Reference

### 5.1 `prepare_selection_data()`

**Purpose:** Preprocesses raw data for selection analysis by standardizing traits, computing relative fitness, and handling missing values. Supports group-level processing for multi-year or multi-site studies.

```r
prepare_selection_data(
  data,                              # Data frame
  fitness_col,                       # Name of fitness column (character)
  trait_cols,                        # Vector of trait column names
  standardize    = TRUE,             # Standardize traits to mean 0, SD 1
  group          = NULL,             # Optional grouping variable (e.g., "year", "site")
  add_relative   = TRUE,             # Add relative fitness column
  na_action      = c("warn", "drop", "none"),
  name_relative  = "relative_fitness"# Name for relative fitness column
)
```

**Returns:** Modified data frame with:
- Standardized traits (if `standardize = TRUE`)
- Relative fitness column (if `add_relative = TRUE`)
- Dropped NA rows (if `na_action = "drop"`)

**Example:**

```{r, message=FALSE, warning=FALSE}
raw_data <- data.frame(
  survival = rbinom(100, 1, 0.6),
  size     = rnorm(100, 10, 2),
  color    = rnorm(100, 5, 1)
)

prepared <- prepare_selection_data(
  data = raw_data,
  fitness_col = "survival",
  trait_cols = c("size", "color"),
  standardize = TRUE,
  add_relative = TRUE,
  na_action = "drop"
)

cat("Rows:", nrow(prepared), "\n")
cat("Trait means after standardization:\n")
print(colMeans(prepared[, c("size", "color")]))
cat("Relative fitness mean:", mean(prepared$relative_fitness), "\n")
head(prepared)
```

### 5.2 `detect_family()`

**Purpose:** Automatically determines the appropriate fitness type and GLM family for p-value calculation based on the distribution of fitness values.

```R
detect_family(y) # y = fitness vector
```

**Detection logic (in order of priority):**
- **Binary**: ≤ 2 unique values, all in {0, 1} → `binomial(logit)`
- **Count**: non-negative integers (> 2 unique values) → `poisson(log)`
- **Proportion**: values in [0, 1] but not binary → `gaussian()` (treated as continuous)
- **Continuous**: all other numeric values → `gaussian()`

**Returns:** List with:
- `$type`: `"binary"`, `"count"`, or `"continuous"`
- `$family`: GLM family object used for p-value calculation
- `$note`: Description of detected fitness type

**Note:** Regardless of detected type, OLS is always used to estimate selection gradients ($\beta$, $\gamma$). The family is used only for obtaining valid p-values via GLM.

**Example:**

```{r, message=FALSE, warning=FALSE}
# Binary fitness (survival)
fitness_binary <- rbinom(50, 1, 0.6)
detect_family(fitness_binary)

# Count fitness (offspring number)
fitness_count <- rpois(50, lambda = 3)
detect_family(fitness_count)

# Continuous fitness (growth rate)
fitness_cont <- rnorm(50, mean = 5, sd = 1)
detect_family(fitness_cont)
```

### 5.3 `selection_differential()`

**Purpose:** Calculates the selection differential ($S$) for a single trait. Supports group-level calculation for multi-year or multi-site studies.

```r
selection_differential(
  data,
  fitness_col,
  trait_col,
  standardized   = TRUE,   # Whether trait is already standardized (mean=0, SD=1)
  use_relative   = TRUE,   # Whether to relativize fitness within function
  group          = NULL,   # Optional grouping variable (e.g., "year", "site")
  return_grouped = FALSE   # If TRUE, return S per group; if FALSE, return weighted mean
)
```

**Formula:**
$$S = \text{Cov}(z, w)$$

When traits are standardized ($\bar{z} = 0$), this simplifies to:
$$S = \frac{1}{n}\sum_{i=1}^n z_i w_i$$

**Returns:**
- `group = NULL`: single numeric value (overall $S$)
- `group` specified + `return_grouped = TRUE`: data frame with $S$ per group
- `group` specified + `return_grouped = FALSE`: weighted mean $S$ across groups

**Example:**
```{r, message=FALSE, warning=FALSE}
S_size <- selection_differential(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_col    = "size",
  standardized = TRUE,
  use_relative = TRUE
)

S_color <- selection_differential(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_col    = "color",
  standardized = TRUE,
  use_relative = TRUE
)

cat("Selection Differentials (S):\n")
cat("Size :", round(S_size, 4), "\n")
cat("Color:", round(S_color, 4), "\n")
```

### 5.4 `analyze_linear_selection()`

**Purpose:** Estimates linear selection gradients ($\boldsymbol{\beta}$) with comprehensive diagnostics. Always uses OLS for gradient estimation; uses logistic GLM for valid p-values when fitness is binary.

```R
analyze_linear_selection(
  data,
  fitness_col,
  trait_cols,
  fitness_type # "binary" or "continuous"
)
```

**Features:**
- Sample size check (warning if n < 10)
- Missing data handling
- Type III ANOVA via `car::Anova()`
- VIF check for multicollinearity (warning if VIF > 5)
- Convergence and separation checks for binary GLM

**Returns:**

For continuous fitness:
- `$model`: `lm` object (OLS — coefficients = $\beta$)
- `$summary`: `summary.lm` object (coefficients, SE, p-values)
- `$anova`: Type III ANOVA table
- `$vif`: Variance inflation factors
- `$fitness_type`: `"continuous"`

For binary fitness:
- `$model$ols`: `lm` object (OLS — coefficients = $\beta$)
- `$model$glm`: `glm` object (logistic — p-values only)
- `$summary$ols`: `summary.lm` (coefficients, SE)
- `$summary$glm`: `summary.glm` (Wald test p-values)
- `$anova`: Type III ANOVA from GLM
- `$vif`: Variance inflation factors from OLS
- `$fitness_type`: `"binary"`

**Example:**

```{r, message=FALSE, warning=FALSE}
linear_results <- analyze_linear_selection(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_cols   = c("size", "color"),
  fitness_type = "continuous"
)

# Coefficients (β) and p-values
summary(linear_results$model)

# Type III ANOVA
linear_results$anova

# VIF
linear_results$vif
```

### 5.5 `analyze_nonlinear_selection()`

**Purpose:** Estimates quadratic ($\gamma_{ii}$) and correlational ($\gamma_{ij}$) selection gradients. Always uses OLS for gradient estimation; uses logistic GLM for valid p-values when fitness is binary.

```r
analyze_nonlinear_selection(
  data,
  fitness_col,
  trait_cols,
  fitness_type
)
```

**Model specification:** Automatically constructs:
- Linear terms: $z_i$
- Quadratic terms: $z_i^2$
- Interaction terms: $z_i \times z_j$ (for $i < j$)

**Diagnostics:**
- Sample size check (warning if n < 20, or n < 2 × number of parameters)
- VIF check for multicollinearity using `type = "predictor"`  (warning if VIF > 5)
- Convergence and separation checks for binary GLM

**Returns:**

For continuous fitness:
- `$model`: `lm` object (OLS — coefficients = $\beta$, $\gamma$)
- `$summary`: `summary.lm` object
- `$anova`: Type III ANOVA table
- `$vif`: Variance inflation factors
- `$fitness_type`: `"continuous"`

For binary fitness:
- `$model$ols`: `lm` object (OLS — coefficients = $\beta$, $\gamma$)
- `$model$glm`: `glm` object (logistic — p-values only)
- `$summary$ols`: `summary.lm`
- `$summary$glm`: `summary.glm`
- `$anova`: Type III ANOVA from GLM
- `$vif`: Variance inflation factors from OLS
- `$fitness_type`: `"binary"`

**Note:** Quadratic coefficients $\gamma_{ii}$ are obtained by multiplying the regression estimate and SE by 2. This is handled automatically by `extract_results.R` — do not multiply manually when using `selection_coefficients()`.

**Example:**
```{r, message=FALSE, warning=FALSE}
nonlinear_results <- analyze_nonlinear_selection(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_cols   = c("size", "color"),
  fitness_type = "continuous"
)

# Raw quadratic coefficient from regression
coef(summary(nonlinear_results$model))["I(size^2)", ]

# gamma_ii = 2 × regression coefficient
gamma_size <- 2 * coef(nonlinear_results$model)["I(size^2)"]
cat("gamma_size:", round(gamma_size, 4), "\n")

# VIF
nonlinear_results$vif
```

### 5.6 `analyze_disruptive_selection()`

**Purpose:** Detects disruptive or stabilizing selection on a single trait using a univariate quadratic model. Supports group-level analysis for multi-year or multi-site studies.

```r
analyze_disruptive_selection(
  data,
  fitness_col,
  trait_col,
  fitness_type   = c("binary", "continuous"),
  standardize    = TRUE,
  group          = NULL,   # Optional grouping variable (e.g., "year", "site")
  return_grouped = FALSE   # If TRUE, return results per group
)
```

**Model:** $w \sim z + z^2$

- OLS always used for coefficient estimates
- For binary fitness: logistic GLM used for valid p-values
- $\gamma = 2 \times \hat{\beta}_{z^2}$, $\text{SE}(\gamma) = 2 \times \text{SE}(\hat{\beta}_{z^2})$

**Returns:** Data frame with columns:
- `Term`: trait name or `trait²`
- `Type`: `"Linear"` or `"Quadratic"`
- `Beta_Coefficient`: $\beta$ (linear) or $\gamma$ (quadratic)
- `Standard_Error`: SE of estimate
- `P_Value`: statistical significance
- `Variance`: $\text{SE}^2$
- `Group`: group identifier (only if `return_grouped = TRUE`)


**Example:**
```{r, message=FALSE, warning=FALSE}
disruptive_test <- analyze_disruptive_selection(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_col    = "size",
  fitness_type = "continuous"
)

print(disruptive_test)
```

### 5.7 `univariate_spline()`

**Purpose:** Fits a flexible nonparametric univariate fitness function using GAMs with penalized regression splines and REML smoothing parameter selection.

```R
univariate_spline(
  data,
  fitness_col,
  trait_col,
  fitness_type = c("binary", "continuous"),
  group        = NULL,  # Optional grouping variable
  k            = NULL
)
```

**Technical details:**
- For binary fitness: binomial GAM with logit link
- For continuous fitness: Gaussian GAM
- `k` is automatically reduced if insufficient unique trait values exist
- Returns predictions with 95% confidence intervals on the response scale

**Returns:** Object of class `univariate_fitness` containing:
- `$model`: fitted GAM object
- `$grid`: data frame with `trait`, `fit`, `lwr`, `upr` columns
- `$fitness_type`: `"binary"` or `"continuous"`
- `$family`: GLM family used
- `$k`: basis dimension used
- `$n_obs`: number of observations used

**Example:**
```{r, message=FALSE, warning=FALSE, cache=FALSE}
# k is selected adaptively by default: min(10, max(5, n_unique - 1))
# Override with k = <value> if needed
spline_fit <- univariate_spline(
  data         = prepared,
  fitness_col  = "relative_fitness",
  trait_col    = "size",
  fitness_type = "continuous"
)

cat("k used:", spline_fit$k, "\n")
cat("Observations:", spline_fit$n_obs, "\n")
head(spline_fit$grid)
```

### 5.8 `plot_univariate_fitness()`

**Purpose:** Visualizes univariate correlated fitness functions with 95% confidence bands.

```r
plot_univariate_fitness(
  uni,                          # Output from univariate_spline()
  trait_col,                    # Trait column name (character)
  title           = NULL,       # Plot title
  show_points     = TRUE,       # Show original data points
  point_alpha     = 0.25,       # Transparency of data points
  ribbon_fill     = NA,         # Fill colour for CI ribbon (NA = no fill)
  ribbon_linetype = "dashed"    # Linetype for CI bounds
)
```

**Returns:** `ggplot2` object with:
- Solid line: predicted fitness
- Dashed lines: 95% confidence bounds
- Optional data points (if `show_points = TRUE` and data stored in object)

**Example:**
```{r, message=FALSE, warning=FALSE, fig.align="center", out.width="80%", fig.cap="Correlated fitness function for body size"}
p <- plot_univariate_fitness(
  uni       = spline_fit,
  trait_col = "size",
  title     = "Correlated Fitness Function: Body Size"
)
print(p)
```


### 5.9 `correlated_fitness_surface()`

**Purpose:** Estimates a 2D correlated fitness surface relating individual fitness to two traits simultaneously, using GAM or thin-plate splines.

```r
correlated_fitness_surface(
  data,
  fitness_col,
  trait_cols,        # Exactly 2 traits
  grid_n       = 60, # Grid resolution (grid_n × grid_n points)
  method       = "auto", # "auto", "gam", or "tps"
  scale_traits = FALSE,  # Deprecated — standardize upstream instead
  group        = NULL,   # Optional grouping variable
  k            = NULL    # Basis dimension; NULL = adaptive
)
```

**Method selection:**
- `"auto"`: GAM for binary fitness, TPS for continuous fitness
- `"gam"`: `mgcv::gam()` with thin-plate spline smooth — works for both binary and continuous
- `"tps"`: `fields::Tps()` — continuous fitness only

**Adaptive k:** When `k = NULL` (default), k is selected as:
$$k = \min\left(30,\; \max\left(10,\; \lfloor\sqrt{n_{\text{unique},1} \times n_{\text{unique},2}}\rfloor\right)\right)$$

**Returns:** List containing:
- `$model`: Fitted model object (`gam` or `Tps`)
- `$grid`: Data frame with trait values and `.fit` column (predicted fitness)
- `$method`: Method used (`"gam"` or `"tps"`)
- `$formula_used`: Formula variant used (GAM only: `"main"`, `"alt1"`, or `"alt2"`)
- `$k_used`: Actual k used for fitting (GAM only)
- `$data_type`: `"binary"` or `"continuous"`
- `$trait_cols`: Trait column names
- `$surface_type`: `"correlated_fitness"`


**Example:**
```{r, message=FALSE, warning=FALSE}
surface <- correlated_fitness_surface(
  data        = prepared,
  grid_n      = 50,
  fitness_col = "relative_fitness",
  trait_cols  = c("size", "color"),
  method      = "gam"
)

cat("Method used:", surface$method, "\n")
cat("k used:", surface$k_used, "\n")

optimum <- surface$grid[which.max(surface$grid$.fit), ]
print(optimum)
```

### 5.10 `plot_correlated_fitness()`

**Purpose:** Visualizes a 2D correlated fitness surface as a filled contour plot.

```r
# Basic version
plot_correlated_fitness(
  tps,                    # Output from correlated_fitness_surface()
  trait_cols,             # Character vector of length 2
  bins         = 12,      # Number of contour bins
  point_alpha  = 0.7,     # Transparency of data points
  show_points  = FALSE,   # Overlay original data points
  show_optimum = TRUE     # Mark optimal trait combination (gold diamond)
)

# Enhanced version — overlays original data coloured by fitness
plot_correlated_fitness_enhanced(
  tps,                    # Output from correlated_fitness_surface()
  trait_cols    = NULL,   # Character vector of length 2 (inferred if NULL)
  original_data = NULL,   # Original data frame for point overlay
  fitness_col   = NULL,   # Fitness column name for point colouring
  bins          = 12,
  point_alpha   = 0.7,
  show_optimum  = TRUE
)
```

**Returns:** `ggplot2` object with filled contour plot. The enhanced version additionally colours data points by fitness value (continuous) or survival outcome.

**Example:**
```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, fig.cap="Correlated fitness surface for size and color"}
# Basic contour plot
p <- plot_correlated_fitness(
  tps        = surface,
  trait_cols = c("size", "color"),
  bins       = 12
)
print(p)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, fig.cap="Correlated fitness surface with original data points"}
# Enhanced version with data overlay
p <- plot_correlated_fitness_enhanced(
  tps           = surface,
  trait_cols    = c("size", "color"),
  original_data = prepared,
  fitness_col   = "relative_fitness",
  bins          = 12
)
print(p)
```

### 5.11 `adaptive_landscape()`

**Purpose:** Calculates the Wrightian adaptive landscape — population mean fitness as a function of population mean phenotype. This is a population-level concept distinct from correlated fitness surfaces (individual-level).

```r
adaptive_landscape(
  data,
  fitness_model,              # Fitted model from correlated_fitness_surface()
  trait_cols,                 # Character vector of length 2
  group_col          = NULL,  # Optional grouping column (e.g., "year", "site")
  population_variance = NULL, # Within-population P matrix; estimated if NULL
  simulation_n       = 1000,  # Individuals simulated per grid point
  grid_n             = 50,    # Grid resolution (grid_n × grid_n)
  custom_range       = NULL   # Optional named list with custom trait ranges
)
```

**Calculation steps:**
1. Create grid of candidate population means ($\bar{z}_1$, $\bar{z}_2$)
2. Estimate within-population variance-covariance matrix $\mathbf{P}$ from data (or use provided)
3. For each grid point, simulate individuals: $\mathbf{z}_j \sim \mathcal{N}(\bar{\mathbf{z}}, \mathbf{P})$
4. Predict individual fitness using the fitted model: $\hat{w}_j = \hat{f}(\mathbf{z}_j)$
5. Compute mean fitness: $\bar{W}(\bar{\mathbf{z}}) = \frac{1}{N}\sum_{j=1}^N \hat{w}_j$
6. Find optimum: $\bar{\mathbf{z}}^* = \arg\max \bar{W}$

**Returns:** Object of class `adaptive_landscape` containing:
- `$grid`: Data frame of population means with `.mean_fit` column
- `$optimum`: Row of `$grid` with highest mean fitness
- `$population_variance`: $\mathbf{P}$ matrix used for simulation
- `$simulation_n`: Number of individuals simulated per grid point
- `$actual_population_means`: Actual group means (if `group_col` provided)
- `$fitness_model_class`: Class of fitted model (`"gam"` or `"Tps"`)
- `$data_summary`: List with n_individuals, trait ranges, standardization check
- `$surface_type`: `"adaptive_landscape"`

**Note:** If the optimum falls at the boundary of the trait grid, it may indicate that the true optimum lies outside the observed trait range. Consider using `custom_range` to extend the grid, or interpreting the result with caution.

**Example:**
```{r, message=FALSE, warning=FALSE}
landscape <- adaptive_landscape(
  data          = prepared,
  fitness_model = surface$model,
  trait_cols    = c("size", "color"),
  grid_n        = 50,
  simulation_n  = 1000
)

# Optimal population mean phenotype
landscape$optimum

# P matrix used
round(landscape$population_variance, 4)
```


### 5.12 `plot_adaptive_landscape.R`

**Purpose:** Visualizes the adaptive landscape as a 2D contour map or 3D perspective plot showing how population mean fitness varies with population mean phenotypes.

```r
# 2D contour plot
plot_adaptive_landscape(
  landscape,                   # Output from adaptive_landscape()
  trait_cols,                  # Two trait column names
  original_data     = NULL,    # Optional: overlay individual data points
  group_col         = NULL,    # Optional: label actual population means by group
  bins              = 12,      # Number of contour bins
  show_optimum      = TRUE,    # Show optimal population mean (gold diamond)
  show_actual_means = TRUE,    # Show actual population means (red points)
  point_alpha       = 0.8      # Transparency of points
)

# 3D perspective plot
plot_adaptive_landscape_3d(
  landscape,                   # Output from adaptive_landscape()
  trait_cols,                  # Two trait column names
  theta         = -30,         # Azimuthal rotation angle
  phi           = 30,          # Colatitude tilt angle
  grid_n        = 200,         # Interpolation resolution (higher = smoother)
  color_palette = NULL         # Colour palette; defaults to viridis::plasma(100)
)
```

**Returns:**
- `plot_adaptive_landscape()`: `ggplot2` object
- `plot_adaptive_landscape_3d()`: base R perspective plot (no `print()` needed)

**Example:**
```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, fig.cap="Adaptive landscape showing mean fitness as a function of population mean size and color"}
p <- plot_adaptive_landscape(
  landscape  = landscape,
  trait_cols = c("size", "color"),
  bins       = 12
)
print(p)
```

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=7, fig.cap="3D adaptive landscape"}
plot_adaptive_landscape_3d(
  landscape  = landscape,
  trait_cols = c("size", "color"),
  theta      = -30,
  phi        = 30,
  grid_n     = 200
)
```

### 5.13 `selection_coefficients()`

**Purpose:** Main wrapper function that runs a complete selection analysis, data preparation, fitness type detection, linear and nonlinear gradient estimation, and coefficient extraction, in a single call.

```r
selection_coefficients(
  data,
  fitness_col,
  trait_cols,
  fitness_type         = c("auto", "binary", "continuous"),
  standardize          = TRUE,
  group                = NULL,   # Optional grouping variable (e.g., "year", "site")
  use_relative_for_fit = TRUE,   # Use relative fitness for continuous data
  return_grouped       = FALSE   # If TRUE and group specified, return results per group
)
```

**Workflow:**
1. Prepares data via `prepare_selection_data()` (standardization, relative fitness)
2. Detects fitness type via `detect_family()` (if `fitness_type = "auto"`)
3. Runs linear selection analysis via `analyze_linear_selection()`
4. Runs nonlinear selection analysis via `analyze_nonlinear_selection()`
5. Extracts and combines all coefficients via `extract_results.R`

**Returns:** Data frame with columns:
- `Term`: coefficient name (e.g., `"size"`, `"size²"`, `"size×color"`)
- `Type`: `"Linear"`, `"Quadratic"`, or `"Correlational"`
- `Beta_Coefficient`: estimated selection gradient ($\beta$ or $\gamma$)
- `Standard_Error`: standard error of estimate
- `P_Value`: statistical significance
- `Variance`: $\text{SE}^2$
- `Group`: group identifier (only if `return_grouped = TRUE`)

**Note:** `prepare_selection_data()` is called automatically — no need to standardize data manually before calling this function.

**Example:**
```{r, message=FALSE, warning=FALSE}
results <- selection_coefficients(
  data         = raw_data,
  fitness_col  = "survival",
  trait_cols   = c("size", "color"),
  fitness_type = "binary",
  standardize  = TRUE
)

print(results)
```

### 5.14 Extraction Functions

Internal functions used by `selection_coefficients()` to parse model outputs into a standardised results table. These are not intended to be called directly by the user.

| Function | Purpose |
|----------|---------|
| `extract_linear_coefficients(trait_cols, results)` | Extracts $\beta$ from linear model |
| `extract_quadratic_coefficients(trait_cols, results)` | Extracts $\gamma_{ii} = 2\hat{\beta}_{ii}$ from nonlinear model |
| `extract_interaction_coefficients(trait_cols, results)` | Extracts $\gamma_{ij}$ from interaction terms |

For binary fitness, all three functions take gradients from OLS and p-values from GLM. For continuous fitness, both come from OLS.

### 5.15 `compare_fitness_surfaces()`

**Purpose:** Quantitatively compares the correlated fitness surface (individual-level) with the adaptive landscape (population-level) by computing optimum locations, their Euclidean distance, and surface correlation.

```r
# Step 1: compute comparison data
compare_fitness_surfaces_data(
  correlated_surface,              # Output from correlated_fitness_surface()
  adaptive_landscape,              # Output from adaptive_landscape()
  trait_cols,
  calculate_correlation = TRUE     # Compute Pearson r between surfaces
)

# Step 2: visualise
plot_fitness_surfaces_comparison(
  comparison_data,                 # Output from compare_fitness_surfaces_data()
  bins        = 10,
  title       = NULL,
  point_alpha = 0.7,
  linewidth   = 0.8,
  show_optima = TRUE
)
```

**Returns from `compare_fitness_surfaces_data()`:**
- `$combined_data`: long-format data frame with both surfaces
- `$summary_stats`: fitness range, mean, SD for each surface
- `$optimum_individual`: trait combination maximising individual fitness
- `$optimum_population`: trait combination maximising population mean fitness
- `$distance_between_optima`: Euclidean distance between the two optima
- `$correlation`: Pearson $r$ between surfaces (if `calculate_correlation = TRUE`)

**Returns from `plot_fitness_surfaces_comparison()`:**
- `$side_by_side`: `patchwork` plot with two contour panels side by side
- `$overlay`: single contour plot overlaying both surfaces

**Example:**

```{r, message=FALSE, warning=FALSE}
comparison <- compare_fitness_surfaces_data(
  correlated_surface = surface,
  adaptive_landscape = landscape,
  trait_cols         = c("size", "color")
)

comparison$summary_stats
comparison$distance_between_optima
comparison$correlation
```

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=5, fig.cap="Overlay comparison of correlated fitness surface and adaptive landscape"}
plots <- plot_fitness_surfaces_comparison(
  comparison_data = comparison,
  bins            = 15
)
print(plots$overlay)
```

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=5, fig.cap="Side-by-side comparison of correlated fitness surface and adaptive landscape"}
print(plots$side_by_side)
```
---

## 6. Interpretation Guide

### 6.1 Statistical Significance

| P-value | Evidence |
|---------|----------|
| < 0.001 | Strong evidence of selection |
| 0.001 – 0.01 | Moderate evidence |
| 0.01 – 0.05 | Weak evidence |
| > 0.05 | No significant selection detected |

**Caution:** Statistical significance depends heavily on sample size. 
With large samples, even biologically trivial gradients may be significant. 
Always interpret p-values alongside effect sizes.

### 6.2 Selection Types

**Linear Selection ($\beta$):**

| $\beta$ sign | Interpretation |
|--------------|----------------|
| Positive (+) | Selection favours larger trait values (positive directional selection) |
| Negative (−) | Selection favours smaller trait values (negative directional selection) |
| ≈ 0 | No directional selection detected |

**Quadratic Selection ($\gamma_{ii}$):**

| $\gamma_{ii}$ sign | Interpretation | Fitness function shape |
|--------------------|----------------|------------------------|
| Negative (−) | Stabilizing selection | ∩-shaped — intermediate values favoured |
| Positive (+) | Disruptive selection | ∪-shaped — extreme values favoured |
| ≈ 0 | No nonlinear selection | Linear fitness function |

**Correlational Selection ($\gamma_{ij}$):**

| $\gamma_{ij}$ sign | Interpretation |
|--------------------|----------------|
| Positive (+) | Selection favours positive trait correlation (both high or both low) |
| Negative (−) | Selection favours negative trait correlation (one high, one low) |
| ≈ 0 | Traits selected independently |

### 6.3 Effect Sizes

Based on the meta-analysis of Kingsolver et al. (2001), which surveyed 
selection gradients across natural populations:

| $|\beta|$ | Effect size |
|-----------|-------------|
| < 0.1 | Weak — below the median observed in nature |
| 0.1 – 0.5 | Moderate — typical range in natural populations |
| > 0.5 | Strong — upper range of selection in nature |

**Reference:** Kingsolver et al. (2001) reported a mean $|\beta| \approx 0.22$ 
and median $|\beta| \approx 0.16$ across studies, suggesting that gradients 
above 0.5 represent unusually strong selection.

### 6.4 Example Interpretations

```
Scenario 1: β_size = 0.25, p = 0.001
  "Body size experiences significant positive directional selection.
   A one-standard-deviation increase in size is associated with a 0.25-unit
   increase in relative fitness, holding other traits constant."

Scenario 2: γ_size = -0.15, p = 0.02
  "Body size shows significant stabilizing selection (γ < 0).
   Individuals with intermediate size have the highest fitness;
   both large and small extremes are disadvantaged."

Scenario 3: γ_size × beak = 0.32, p = 0.004
  "Significant positive correlational selection between size and beak depth.
   Selection favours individuals that are either both large-billed and large-bodied,
   or both small-billed and small-bodied."
```
---

## 7. Troubleshooting and Common Issues

### 7.1 Sample Size Recommendations

| Analysis | Minimum n | Recommended n |
|----------|-----------|---------------|
| Linear selection (continuous) | 10 | > 30 |
| Linear selection (binary) | 20 per group | > 50 per group |
| Nonlinear selection | 20 | > 100 |
| Bivariate GAM / TPS surface | 30 | > 200 |
| Adaptive landscape (simulation) | — | > 100 (for stable $\mathbf{P}$ estimate) |

### 7.2 Convergence Issues in Binary Models

**Symptoms:**
- Very large coefficients (> 10) — detected automatically by the framework
- Extremely large standard errors
- Warning: `"GLM did not converge"`
- Warning: `"Possible complete separation detected"`

**Solutions:**
```r
# 1. Check for complete separation
table(data$fitness, cut(data$trait, 5))

# 2. Use Firth's penalized likelihood
if (!require("logistf")) install.packages("logistf")
library(logistf)
fit <- logistf(fitness ~ trait, data = data)

# 3. Collect more data
# 4. Simplify model (fewer traits)
```

### 7.3 Missing Data Handling

```r
# Check missing data pattern
apply(data, 2, function(x) sum(is.na(x)))

# Option 1: Drop rows with missing (if < 5%)
prepared <- prepare_selection_data(..., na_action = "drop")

# Option 2: Use na_action = "warn" to investigate without dropping
prepared <- prepare_selection_data(..., na_action = "warn")

# Option 3: Impute missing values before passing to the framework
library(mice)
imp          <- mice(data, m = 5)
data_complete <- complete(imp, 1)
prepared     <- prepare_selection_data(data_complete, ...)
```

### 7.4 Multicollinearity

High correlations between traits can inflate standard errors. The framework 
automatically checks VIF and warns if VIF > 5:

```r
# Check pairwise correlations before analysis
cor(data[, trait_cols])

# Retrieve VIF from linear model results
linear_results$vif

# Retrieve VIF from nonlinear model results
nonlinear_results$vif
```

**Interpreting VIF:**

| VIF | Interpretation |
|-----|----------------|
| 1 – 5 | Acceptable |
| 5 – 10 | Moderate collinearity — interpret SE with caution |
| > 10 | Severe collinearity — results unreliable |

**Solutions:**
- Remove the more collinear trait and run separate univariate analyses
- Use principal component analysis (PCA) to create orthogonal trait axes
- Collect more data to improve estimation stability

### 7.5 Adaptive k Warning

If `univariate_spline()` or `correlated_fitness_surface()` warns that k has 
been reduced, it means the subset has fewer unique trait values than the 
adaptive default:

```r
# Override adaptive k manually if needed
spline_fit <- univariate_spline(..., k = 5)
surface    <- correlated_fitness_surface(..., k = 15)
```

### 7.6 Grid Size Mismatch in Surface Comparison

`compare_fitness_surfaces_data()` requires both surfaces to have the same 
`grid_n` to compute surface correlation. If they differ, correlation returns 
`NULL`:

```r
# Use matching grid_n for both surfaces
surface   <- correlated_fitness_surface(..., grid_n = 50)
landscape <- adaptive_landscape(..., grid_n = 50)
```

---

## 8. References

1. Lande, R., & Arnold, S. J. (1983). The measurement of selection on correlated characters. *Evolution*, 37(6), 1210-1226.

2. Arnold, S. J., & Wade, M. J. (1984). On the measurement of natural and sexual selection: Theory. *Evolution*, 38(4), 709-719.

3. Brodie, E. D., Moore, A. J., & Janzen, F. J. (1995). Visualizing and quantifying natural selection. *Trends in Ecology & Evolution*, 10(8), 313-318.

4. Schluter, D. (1988). Estimating the form of natural selection on a quantitative trait. *Evolution*, 42(5), 849-861.

5. Wood, S. N. (2017). Generalized Additive Models: An Introduction with R (2nd ed.). Chapman and Hall/CRC.

6. Kingsolver, J. G., et al. (2001). The strength of phenotypic selection in natural populations. *The American Naturalist*, 157(3), 245-261.

7. Hereford, J., Hansen, T. F., & Houle, D. (2004). Comparing strengths of directional selection: How strong is strong? *Evolution*, 58(10), 2133-2143.

